# Bayesian Inference {#sec-bayes-inference}

```{python}
import matplotlib.pyplot as plt

plot_bg = "#f8f8f8"

plt.rcParams["figure.facecolor"] = plot_bg
plt.rcParams["axes.facecolor"]   = plot_bg

```

## Introduction to the Bayesian approach

As discussed in the course introduction, **Bayesian inference** takes a different perspective from the frequentist approach:

* In the frequentist view, the parameter $\theta$ is **fixed but unknown**.
* In the Bayesian view, the parameter $\theta$ is treated as a **random variable**.

The randomness here does not mean that the parameter is *physically fluctuating* — instead, it reflects our **uncertainty** about its value.

### Why is this useful?

A **key advantage** of Bayesian inference is that it allows us to **incorporate prior information**[^prior] about the parameter.

::: {.definition #def-prior}
### Prior Distribution.
Our uncertainty about $\theta$ is represented by a probability distribution, called the **prior distribution**:

$$
\theta \sim \pi(\theta).
$$

Here $\pi(\theta)$ denotes the prior density (PDF or PMF).
:::

::: {.worked-example #exm-prior}
Suppose $\theta$ is the average exam score of students in a module. From past experience, the average score tends to be around 60, with most cohorts lying within about 10 marks of this.

We could encode this information as a **Normal prior**:

$$
\theta \sim \mathcal{N}(\mu=60,\;\sigma^2=10^2).
$$

This prior expresses our belief *before seeing the new data*.
:::



::: {.panel-tabset}

### Python

```{python}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
#| fig-align: center

import matplotlib.pyplot as plt
import numpy as np
import math

def normal_pdf(theta, mean, sd):
    const = 1 / (sd * (2 * math.pi) ** 0.5)
    exp_term = np.exp(-((theta - mean) / sd) ** 2 / 2)
    return const * exp_term

mean = 100.0
sd   = 10.0

theta_grid = np.linspace(mean - 3*sd, mean + 3*sd, 200)
pdf_values = normal_pdf(theta_grid, mean=mean, sd=sd)

plt.figure(figsize=(7,4))
plt.plot(theta_grid, pdf_values)
plt.xlabel(r"$\theta$")
plt.ylabel("Density")
plt.title("Normal PDF: mean = 100, sd = 10")
plt.tight_layout()

```

### R

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
#| fig-align: center

mean <- 100
sd   <- 10

theta_grid <- seq(mean - 3*sd, mean + 3*sd, length.out = 200)
pdf_values <- dnorm(theta_grid, mean = mean, sd = sd)

plot(theta_grid, pdf_values, type = "l",
     xlab = expression(theta),
     ylab = "Density",
     main = "Normal PDF: mean = 100, sd = 10")
```
:::


## Bayes’ Theorem

Bayesian inference takes its name from **Bayes’ theorem**, which tells us how to update our beliefs about $\theta$ after seeing data.

Recall from probability that for events $A$ and $B$:

$$
\mathrm{Pr}(A \mid B) = \frac{\mathrm{Pr}(B \mid A)\,\mathrm{Pr}(A)}{\mathrm{Pr}(B)}.
$$

To generalise this idea to **random variables and densities**, we use conditional densities:

$$
f_{X \mid Y}(x \mid y) = \frac{f_{XY}(x,y)}{f_Y(y)}.
$$

::: {.remark #rem-notation}
**Notation.**
We will usually write $f(x \mid y)$ rather than $f_{X \mid Y}(x \mid y)$ to keep things simple. Both mean the same thing.
:::

Using this, we obtain Bayes’ theorem for densities.

::: {.thm #thm-bayes}
### Bayes’ Theorem for Densities.
For random variables $X$ and $Y$:

$$
f(x \mid y) = \frac{f(y \mid x)\,f(x)}{f(y)}.
$$

:::

::: {.prf #prf-bayes}
By the definition of conditional density:

$$
f(x \mid y) = \frac{f(x,y)}{f(y)}, 
\qquad
f(y \mid x) = \frac{f(x,y)}{f(x)}.
$$

Rearranging the second expression gives $f(x,y) = f(y \mid x) f(x)$. Substituting this into the first gives

$$
f(x \mid y) = \frac{f(y \mid x) f(x)}{f(y)}.
$$
:::

---

## Bayes’ Theorem in Statistics

In Bayesian inference, we apply Bayes’ theorem with:

* $x = \theta$ (the parameter, treated as random),
* $y = \underline{x}$ (the observed data).

This gives:

$$
\pi(\theta \mid \underline{x}) 
= \frac{f(\underline{x} \mid \theta)\,\pi(\theta)}{f(\underline{x})}.
$$

::: {.definition #def-posterior}
### Posterior Distribution.

$$
\pi(\theta \mid \underline{x}) 
= \frac{f(\underline{x} \mid \theta)\,\pi(\theta)}{f(\underline{x})}.
$$

The **posterior** combines two ingredients:

* The **likelihood** $f(\underline{x} \mid \theta)$ (information from the data).
* The **prior** $\pi(\theta)$ (information before seeing the data).

The denominator $f(\underline{x})$ is a normalising constant ensuring the posterior integrates to 1.
:::

::: {.callout-important appearance="minimal"}
## ✨ Big picture:
Bayesian inference updates prior beliefs about parameters using the likelihood from the data, producing the posterior distribution which represents *updated uncertainty*.
:::

Let's start with a simple example to see how this works:

::: {.example #exm-biased-coin}
## Biased Coin

We consider a possibly biased coin with parameter $\theta = \Pr(\text{Head})$.
Before seeing data, we express *no preference* over $\theta\in(0,1)$ via a uniform prior:
$$
\theta \sim \mathrm{Uniform}(0,1) \equiv \mathrm{Beta}(1,1), \qquad \pi(\theta)=1 \ \text{for } 0<\theta<1.
$$
Hence $\mathrm{E}[\theta]=0.5$.

We toss the coin $n=5$ times and observe $x=1$ head. Find the **posterior** $\pi(\theta\mid x)$.
:::

::: {.solution #sol-biased-coin}
Let $X$ be the number of heads in $5$ tosses. Assuming that each coin throw is *independent and identically distributed*, $X$ is Binomial[^binomial]:
$$
    X \mid \theta \sim \textrm{Bin}(5,\theta).
$$
Note that there is only a **single observation** here: one experiment is "tossing the coin $5$ times and seeing how many come up heads". The probability of observing $X=1$ is
$$
    f(x = 1 \mid \theta) = 5\theta(1-\theta)^4
$$
If we plot this:

```{python}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
#| fig-align: center

import matplotlib.pyplot as plt
import numpy as np
import math

def binom_lik(theta):
    return 5 * theta * (1 - theta) ** 4


theta_grid = np.linspace(0, 1, 200)
binom_lik_vals = binom_lik(theta_grid)

plt.figure(figsize=(5,3))
plt.plot(theta_grid, binom_lik_vals)
plt.xlabel(r"$\theta$")
plt.ylabel("Likelihood")
plt.title("Binomial Likelihood (with $n=5$ and $X=1$)")
plt.tight_layout()
plt.show()  
```

we see that this favours $\theta$ around $0.2$. In fact, in @exm-binom-mle-single, we saw that the MLE is $\hat{\theta} = 0.2$.

Our prior for $\theta$ was $\pi(\theta) = 1$, for $0 < \theta < 1$. To update our beliefs, by conditioning on the single observation $x=2$, we use <span class="xref-no-num">[Bayes Theorem @def-posterior]</span>:
$$
    \pi(\theta|x=1)=\frac{\pi(\theta)f(x=1|\theta)}{f(x=1)} = \frac{1 \times 5\theta(1-\theta)^4}{f(x=1)}, \quad 0 < \theta < 1.
$$
To compute the *denominator*, $f(x=1)$, we use the <span class="xref-no-num">[law of total probability @thm-total-probability-pdfs]</span>:
$$
\begin{align*}
f(x=1) &=\int_\Theta\pi(\theta)f(x=1\, |\,\theta)\,d\theta = \int_0^1 1\times 5\theta(1-\theta)^4\,d\theta \\
&=\int_0^1 \theta\times 5(1-\theta)^4\,d\theta =\left[-(1-\theta)^5\,\theta\right]^1_0
+\int_0^1 (1-\theta)^5\,d\theta \\
&=0 + \left[-\frac{(1-\theta)^6}{6}\right]^1_0 =\frac{1}{6}. 
\end{align*}
$$

Therefore, the posterior density is (for $0<\theta<1$):
$$
\begin{align*}
\pi(\theta|x=1)&=\frac{\pi(\theta)f(x=1|\theta)}{f(x=1)} =\frac{5\theta(1-\theta)^4}{1/6}\\
&=30\,\theta(1-\theta)^4 =\frac{\theta(1-\theta)^4}{\mathrm{B}(2,5)},\quad\quad 0<\theta<1,
\end{align*}
$$
and so the posterior distribution is $\theta|x=1\sim \textrm{Beta}(2,5)$. This distribution has its mode at
$\theta=0.2$, and mean at $\mathrm{E}[\theta|x=1]=2/7=0.286$.

:::


The main difficulty in calculating the posterior distribution was in obtaining the $f(x)$ term. However, in many cases we can recognise the posterior distribution without the need to calculate this constant term (constant with respect to $\theta$). In this example, we can calculate the posterior distribution as
$$
\begin{align*}
\pi(\theta|\underline{x})&\propto\pi(\theta)f(x=1|\theta) \\
&\propto 1\times 5\theta(1-\theta)^4,\quad\quad 0<\theta<1  \\
&=k\theta(1-\theta)^4,\quad\quad 0<\theta<1.
\end{align*}
$$

As $\theta$ is a continuous quantity, what we would like to know is what continuous distribution defined on $(0,1)$ has a probability density function which takes the form $k\theta^{g-1}(1-\theta)^{h-1}$. The answer is the $\textrm{Beta}(g,h)$ distribution. Therefore, choosing $g$ and $h$ appropriately, we can see that the posterior distribution is $\theta|x=1\sim \textrm{Beta}(2,5)$.

**Summary:** 

It is possible that we have a biased coin. If we suppose that all values of $\theta=\text{Pr(Head)}$ are equally likely and then observe 1 head out of 5, then the most likely value of $\theta$ is 0.2 --- the same as the most likely value from the data alone (not surprising!). However, on average, we would expect $\theta$ to be around 0.286. Uncertainty about $\theta$ has changed from a (prior) standard deviation of 0.289 to a (posterior) standard deviation of 0.160. The changes in our beliefs about $\theta$ are more fully described by the prior and posterior distributions shown in @fig-beta-posterior.

---

```{ojs}

// Domain and grid
mDomain = [0, 1]
pdfXs = d3.scaleLinear().domain(mDomain).ticks(200)  // 201 points including 0 and 1

// PDFs (closed-form)
function priorPDF(x)     { return (x > 0 && x < 1) ? 1 : 0; }
function posteriorPDF(x) { return (x > 0 && x < 1) ? 30 * x * (1 - x) ** 4 : 0; }

// Data (use the scalar x from map!)
prior = pdfXs.map(x => ({ x, y: priorPDF(x), which: "Prior  Beta(1,1)" }))
post  = pdfXs.map(x => ({ x, y: posteriorPDF(x), which: "Posterior  Beta(2,5)" }))
```

```{ojs}
//| label: fig-beta-posterior
//| fig-cap: Prior and posterior distributions for the coin bias parameter $\theta=\Pr(\text{Head})$ after observing 1 head in 5 tosses. The posterior is more concentrated, reflecting reduced uncertainty about $\theta$ compared to the uniform prior
//| fig-align: center

bodyFont = getComputedStyle(document.body).fontFamily
// Figure
viewof fig_beta = Plot.plot({
  width: 640,
  height: 320,
  marginLeft: 48,
  marginBottom: 40,
  x: { label: "θ", domain: [0, 1] },
  y: { label: "Density" },
  style: { background: "var(--plot-panel-bg)", 
           fontSize: 14,
           fontFamily: bodyFont,  },
  marks: [
    Plot.ruleY([0], { stroke: "var(--border)", strokeOpacity: 0.6 }),
    Plot.lineY(prior, { x: "x", y: "y", stroke: "var(--brand-red)",  strokeWidth: 2, strokeOpacity: 0.85 }),
    Plot.areaY(prior, { x: "x", y: "y", fill: "var(--brand-red)", fillOpacity: 0.05 }),
    Plot.lineY(post,  { x: "x", y: "y", stroke: "var(--brand-teal)", strokeWidth: 2 }),
    Plot.areaY(post,  { x: "x", y: "y", fill: "var(--brand-teal)", fillOpacity: 0.2 }),
    Plot.text(
      [
        { x: 0.8, y: priorPDF(0.8),     label: "Prior" },
        { x: 0.35, y: posteriorPDF(0.3), label: "Posterior" }
      ],
      { x: "x", y: "y", text: "label", dy: -8 }
    )
  ]
})

```

---

::: {#exm-music-expert}
Consider an experiment to determine how good a music expert is at distinguishing between pages from Haydn and Mozart scores. Let $\theta=\mathrm{Pr}(\text{correct choice})$. Suppose that, before conducting the experiment, we have been told that the expert is very competent. In fact, it is suggested that we should have a prior distribution which has a mode around $\theta=0.95$ and for which $\mathrm{Pr}(\theta<0.8)$ is very small. We choose $\theta\sim \textrm{Beta}(77,5)$, with probability density function
$$
\pi(\theta)=128107980\,\theta^{76}(1-\theta)^4,\quad\quad 0<\theta<1.
$$
A graph of this prior density is given as follows:

::: {.panel-tabset}

### Python


```{python}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
#| fig-align: center

import numpy as np
import math
import matplotlib.pyplot as plt

# Beta(α, β) with α=77, β=5  →  π(θ) ∝ θ^(76) (1-θ)^(4)
alpha, beta = 77, 5

# B(α,β) = Γ(α)Γ(β)/Γ(α+β)  ⇒  1/B = exp(lgamma(α+β) - lgamma(α) - lgamma(β))
# Note the lgamma function computes the logarithm of the Gamma function.
beta_const_inv = math.exp(math.lgamma(alpha + beta) - math.lgamma(alpha) - math.lgamma(beta))

def beta_pdf(theta):
    return beta_const_inv * np.power(theta, alpha - 1) * np.power(1 - theta, beta - 1)

theta_grid = np.linspace(0.0, 1.0, 500)
pdf_values = beta_pdf(theta_grid)

plt.figure(figsize=(7,4))
plt.plot(theta_grid, pdf_values)
plt.xlabel(r"$\theta$")
plt.ylabel("Density")
plt.title(rf"Beta PDF: $\alpha={alpha}$, $\beta={beta}$")
# plt.xlim(0, 1)
plt.tight_layout()
plt.show()
```

### R

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
#| fig-align: center

alpha <- 77
beta  <- 5

# Normalising constant via log-gamma (works like Python's lgamma)
beta_const_inv <- exp(lgamma(alpha + beta) - lgamma(alpha) - lgamma(beta))

beta_pdf <- function(theta) {
  beta_const_inv * theta^(alpha - 1) * (1 - theta)^(beta - 1)
}

theta_grid <- seq(0, 1, length.out = 500)
pdf_values <- beta_pdf(theta_grid)

plot(theta_grid, pdf_values, type = "l",
     xlab = expression(theta), ylab = "Density",
     main = bquote("Beta PDF: " ~ alpha == .(alpha) ~ "," ~ beta == .(beta)))
```

:::


:::


[^prior]:
  **Prior distribution.** In Bayesian inference, the *prior* represents our beliefs about the parameter $\theta$ **before seeing any data**. It is written as a probability distribution $\pi(\theta)$, which can be based on previous studies, expert knowledge, or simply chosen to reflect uncertainty (e.g. a flat prior).

[^binomial]:
  **Binomial distribution.** If we assume each coin throw $C_i$ is an i.i.d. $\textrm{Bernoulli}(\theta)$ trial with a $1$ representing a heads landing, and a $0$ representing a tails, the total number of heads out of $5$ is $X = C_1 + C_2 + C_3 + C_4 + C_5$. Therefore:
    $$
        X \mid \theta \sim \textrm{Bin}(5,\theta).
    $$


## Bayesian vs. Frequentist (again)

